sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Ref apps

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}, Appendix 1).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

@@

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}, Appendix 1).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix 1).
The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].

@@

\begin{landscape}
\begin{figure}[p]
\centering 
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Appendix 1 for a full description of the model.
Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} [@wood2017a], \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`.
We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package.
We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument).
<!-- We could instead model catch weight as the response with an offset for log(area). -->

@@

<!-- Setting `silent = FALSE` would indicate optimization progress, which is helpful for larger models. -->
We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.
<!-- (omitted for space). -->

@@

We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
Here we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as an AR(1) process (`time_varying_type = "ar1"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.

@@

```{r fit-tv, eval=TRUE, echo=TRUE, cache=TRUE, results='hide'}
fit_spatiotemporal <- sdmTMB(
  density ~ 0 + s(depth, k = 5), 
  time_varying = ~ 1,
  family = tweedie(link = "log"), 
  data = pcod,
  mesh = mesh,
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  time_varying_type = "ar1",
  anisotropy = TRUE,
  silent = FALSE
)
```

@@

This model took approximately `r round(as.numeric(t2 - t1), 0)` seconds to fit on our laptop. 
We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix 2).

@@

We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig.\ \ref{fig:owl-nao}, Appendix 3).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.

@@

We ran a simple speed comparison between \pkg{sdmTMB}, \pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix 6).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
\pkg{mgcv} was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface and speed.
We think users familiar with `stats::glm()`, \pkg{lme4}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
Users familiar with \pkg{INLA} will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models;
alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages \pkg{sdmTMB} and \pkg{VAST} were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->

>>>>

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

@@

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix \ref{app:model}).
The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].

@@

\begin{landscape}
\begin{figure}[p]
\centering 
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Appendix \ref{app:model} for a full description of the model.
Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} [@wood2017a], \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`.
We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package.
We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument).

@@

We can get a summary of the fit with the `print()` or `summary()` methods or extract parameter estimates as a data frame with the `tidy()` method.

@@

We could extend this spatial model to be a spatiotemporal one simply by supplying the year column name to the `time` argument (`time = "year"`) and specifying how we want the spatiotemporal random fields to be structured.
Here we will keep the spatial field (`spatial = "on"`), structure the spatiotemporal random fields as independent (`spatiotemporal = "iid"`), include a time-varying intercept (`~ 1`) fit as a random walk process (`time_varying_type = "rw"`), include anisotropy to allow spatial correlation to decay at different rates up the coast vs.\ away from the coast (`anisotropy = TRUE`), and use `silent = FALSE` to see model fitting progress.

@@

```{r fit-tv, eval=TRUE, echo=TRUE, cache=TRUE, results='hide'}
fit_spatiotemporal <- sdmTMB(
  density ~ 0 + s(depth, k = 5), 
  time_varying = ~ 1,
  family = tweedie(link = "log"), 
  data = pcod,
  mesh = mesh,
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  time_varying_type = "rw",
  anisotropy = TRUE,
  silent = FALSE
)
```

@@

<!-- This model took approximately `r round(as.numeric(t2 - t1), 0)` seconds to fit on our laptop.  -->
We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).

@@

We found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig.\ \ref{fig:owl-nao}).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
We provide an extended version of this example in Appendix \ref{app:owls}.

@@

We ran a simple speed comparison between \pkg{sdmTMB}, \pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
\pkg{mgcv} was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
We think users familiar with `stats::glm()`, \pkg{lme4}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
Users familiar with \pkg{INLA} will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models;
alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages \pkg{sdmTMB} and \pkg{VAST} were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->
